---
title: "Cost-effectiveness with magentabook and greenbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cost-effectiveness with magentabook and greenbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(magentabook)
greenbook_available <- requireNamespace("greenbook", quietly = TRUE)
```

The Green Book covers *appraisal* (decide whether to fund a policy). The Magenta
Book covers *evaluation* (learn whether the funded policy worked). The two
packages compose cleanly: `greenbook` discounts and rebases cashflows;
`magentabook` turns the resulting present values into cost-effectiveness
ratios and net benefits.

This vignette compares two delivery options for a hypothetical GBP 1.5m health
intervention: an enhanced model that costs more upfront but is expected to
deliver more QALYs.

## Step 1: build the cashflows

```{r}
# 5-year horizon, real GBP, 2026 prices
years <- 0:4

# Status quo: GBP 1m capex in year 0, then GBP 100k opex per year
cost_a <- c(1e6, 1e5, 1e5, 1e5, 1e5)

# Enhanced: GBP 1.5m capex, GBP 150k opex
cost_b <- c(1.5e6, 1.5e5, 1.5e5, 1.5e5, 1.5e5)

# QALY profile: enhanced delivers more in steady-state
qaly_a <- c(0, 80, 80, 80, 80)
qaly_b <- c(0, 100, 110, 115, 115)
```

## Step 2: discount with greenbook

If `greenbook` is available, use the kinked STPR (3.5 percent over the first
30 years for a non-health appraisal; or 1.5 percent under the health schedule):

```{r, eval = greenbook_available}
gb_npv_cost_a <- greenbook::gb_npv(-cost_a, schedule = "health")
gb_npv_cost_b <- greenbook::gb_npv(-cost_b, schedule = "health")

gb_npv_qaly_a <- greenbook::gb_npv(qaly_a, schedule = "health")
gb_npv_qaly_b <- greenbook::gb_npv(qaly_b, schedule = "health")

# Convert to absolute (positive) PV costs
pv_cost_a <- abs(gb_npv_cost_a)
pv_cost_b <- abs(gb_npv_cost_b)

c(pv_cost_a = pv_cost_a, pv_cost_b = pv_cost_b)
c(pv_qaly_a = gb_npv_qaly_a, pv_qaly_b = gb_npv_qaly_b)
```

```{r, include = FALSE}
# Fallback values for when greenbook is not installed (vignette must still build)
if (!greenbook_available) {
  # Approximate health-schedule discounting at r = 0.015
  df <- 1 / (1 + 0.015)^(0:4)
  pv_cost_a   <- sum(cost_a * df)
  pv_cost_b   <- sum(cost_b * df)
  gb_npv_qaly_a <- sum(qaly_a * df)
  gb_npv_qaly_b <- sum(qaly_b * df)
}
```

## Step 3: cost-effectiveness with magentabook

Plain cost per QALY for each option:

```{r}
cea_a <- mb_cea(cost = pv_cost_a, effect = gb_npv_qaly_a, label = "Status quo")
cea_b <- mb_cea(cost = pv_cost_b, effect = gb_npv_qaly_b, label = "Enhanced")
cea_a
cea_b
```

Incremental analysis (B vs A):

```{r}
icer <- mb_icer(
  cost_a   = pv_cost_a, effect_a = gb_npv_qaly_a,
  cost_b   = pv_cost_b, effect_b = gb_npv_qaly_b,
  label_a  = "Status quo", label_b  = "Enhanced"
)
icer
```

The dominance flag tells us which quadrant of the cost-effectiveness plane the
enhanced option sits in.

## Step 4: net benefit at standard willingness-to-pay thresholds

NICE's reference WTP for a QALY is GBP 20k-30k. The cross-government Magenta
Book equivalent is GBP 70k per QALY. Compute incremental net benefit at each:

```{r}
sapply(
  c(NICE_low = 20000, NICE_high = 30000, MB_central = 70000),
  function(wtp) mb_inb(icer$delta_cost, icer$delta_effect, wtp)
)
```

Positive INB means the option is cost-effective at that WTP.

## Step 5: probabilistic sensitivity

Real evaluations carry uncertainty in both costs and effects. Suppose a
probabilistic sensitivity analysis (e.g. Monte-Carlo over an underlying
trial's posterior) gives sampled draws of the incremental cost and
incremental effect:

```{r}
set.seed(20260427)
n_draws <- 5000
delta_cost   <- rnorm(n_draws, mean = icer$delta_cost,   sd = 1e5)
delta_effect <- rnorm(n_draws, mean = icer$delta_effect, sd = 30)

ceac <- mb_ceac(
  delta_cost, delta_effect,
  wtp_grid = seq(0, 100000, by = 5000)
)
ceac
```

Each row is the probability that the enhanced option is cost-effective at the
corresponding WTP. The CEAC is the standard cost-effectiveness uncertainty
visualisation; pipe `ceac$wtp` and `ceac$prob_cost_effective` to your
plotting library of choice.

## Step 6: report

```{r}
sms_b <- mb_sms_rate(
  level = 5, study = "Pilot RCT of the enhanced option",
  design = "Cluster RCT, 30 GP practices",
  notes = "Power calculation per mb_sample_size()"
)

conf <- mb_confidence(
  rating                 = "high",
  question               = "Does the enhanced option deliver more QALYs",
  evidence_strength      = "Single Level 5 cluster RCT plus modelled extrapolation",
  methodological_quality = "Strong: randomisation worked, follow-up rate > 90%",
  generalisability       = "Tested in a representative sample of UK GP practices",
  rationale              = "RCT plus consistent observational evidence"
)

report <- mb_evaluation_report(
  toc = mb_theory_of_change(
    inputs = "GBP 1.5m capex + GBP 150k opex p.a.",
    activities = "Enhanced clinical pathway",
    outputs = "More patients treated to standard",
    outcomes = "Higher QALYs gained per patient",
    impact = "Improved population health"
  ),
  sms = sms_b,
  confidence = conf,
  cea = list(cea_a, cea_b, icer),
  name = "Enhanced clinical pathway evaluation"
)
report
```

Export the bundle to LaTeX for inclusion in a one-page committee paper:

```{r}
cat(mb_to_latex(report, caption = "Enhanced pathway evaluation summary"))
```

## Why this composition matters

By the time a Magenta Book evaluation is asked to compute a cost-effectiveness
ratio, the cashflows are usually nominal and unaligned with the appraisal-stage
price base. Without `greenbook`, the analyst hand-discounts in a spreadsheet
and the audit trail is fragile. With both packages loaded, the appraisal-stage
discount factors and the evaluation-stage cost-effectiveness primitives draw
from the same vintage-tagged parameter tables, and the entire chain is testable
R code.
